Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
minima

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of the Double Pushout (DPO)
  Approach to rewriting in adhesive categories, a very general setting
  for rewriting.  More precisely, we focus on left-linear rules, which
  not only consume, preserve and generate entities, as linear rules
  do, but may ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails,
% : the existence of a consistent permutation is only necessary but
% not sufficient
hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

\begin{notation}
  We stipulate some notational conventions that will be used
  throughout the paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects: so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

@@

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property that $\mathcal{M}$-adhesive categories enjoy is given
by the so-called \emph{Van Kampen condition}
\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We
will recall it and examine some of its consequences. First of all, we
need to recall some terminology and facts regarding subclasses of
$\mor(\X)$.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the ``Double-Pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$, such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ is the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

@@

  \noindent
  \parbox{10cm}{\hspace{15pt}Given two objects $G$ and $H$ and a rule
    $\rho=(l,r)$ in $\R$, a \emph{direct derivation $\mathscr{D}$ from
      $G$ to $H$ applying the rule $\rho$} is a diagram as the one
    aside, in which both squares are pushouts. The arrow $m$ is called
    the \emph{match} of the derivation, while $h$ is its
    \emph{back-match}.  We will denote a direct derivation $\dder{D}$
    between $G$ and $H$ as $\dder{D}\colon G\Mapsto H$. }
  \parbox{3cm}{$\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}&
      H}$}
\end{definition}

@@

Derivations naturally compose: given $\der{D}: G \Mapsto H$ and $\der{E}: H \Mapsto F$ we can consider their composition $\der{D} \cdot \der{E}: G \to F$, defined in the obvious way.

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, a fundamental ingredient of the rewriting
theory, which comes into play when viewing a sequence of steps as a
concurrent computation. We observe that moving from linear to
left-linear rules leads to the failure of various basic properties of
the classical notion of independence used in the linear setting and we
single out a framework in which these can be re-established, possibly
in a weakened form.

@@

In the DPO approach the canonical notion of independence is that of sequential independence.
%recalled below.

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

Switch equivalence is now defined as the equivalence relating derivations which can be obatained one from another by a sequence of switches. Moreover, intermediate graphs can be taken up to isomorphism according to abstraction equivalence.

@@

The fact that sequential indepedence implies switchability always
holds for linear rules (see \Cref{prop:equi}). The result is so
indispensable that it is not even stated, in the sense that
in the literature switchability is not introduced axiomatically as in
Definition~\ref{def:switch}, but it is based on the explicit
construction of a switch.

@@

  Take now the set $S:=\{a,b,c\}$ and endow it with the smallest order
  such that $a\leq b$ and $a\leq c$.  Take also the set $\mathbb{N}$
  ordered by $\geq$, so that $0$ is its maximum.  Let $\X$ be the
  category associated to the order
  $(S, \leq)\oplus (\mathbb{N}, \geq)$, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. Notice that in this category the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a pushout
  because $b$ and $c$ do not have a supremum.  Consider a rewriting
  system whose set of rules $\R$ contains the following two rules
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ of length $2$ below
  \[
    \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
      \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
      \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
      \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback object (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

\begin{remark}
  By \Cref{thm:church} the existence of a strong independence pair
  entails switchability which, in turn, entails weak sequential
  independence by construction. Strong Enforcing rewriting systems are exactly
  those rewriting systems in which these three notions coincide.
\end{remark}
%}

As a sanity check note that a source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

>>>>

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of the Double Pushout (DPO)
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  More precisely, we focus on left-linear rules, which
  not only consume, preserve and generate entities, as linear rules
  do, but may ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails,
% : the existence of a consistent permutation is only necessary but
% not sufficient
hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the two properties above hold.
We show that much of the theory can be retained: consistency
of switching still holds, while globality of independence  in a
weaker form only, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

\begin{notation}
  We stipulate some notational conventions that will be used
  throughout the paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects, so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

@@

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property that $\mathcal{M}$-adhesive categories enjoy is given
by the \emph{Van Kampen condition}
\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We
will recall it and examine some of its consequences. First of all, we
need to introduce some terminology and facts regarding subclasses of
$\mor(\X)$.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
approach (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ is the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

@@

  \vspace{.1cm}
  \noindent
  \parbox{10cm}{\hspace{15pt}Given two objects $G$ and $H$ and a rule
    $\rho=(l,r)$ in $\R$, a \emph{direct derivation $\mathscr{D}$ from
      $G$ to $H$ applying the rule $\rho$} is a diagram as the one
    aside, in which both squares are pushouts. The arrow $m$ is called
    the \emph{match} of the derivation, while $h$ is its
    \emph{back-match}.  We will denote a direct derivation $\dder{D}$
    between $G$ and $H$ as $\dder{D}\colon G\Mapsto H$. }
  \parbox{3cm}{$\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}&
      H}$}
\end{definition}

@@

Derivations naturally compose: given $\der{D}: G \Mapsto H$ and $\der{E}: H \Mapsto F$, we can consider their composition $\der{D} \cdot \der{E}: G \to F$, defined in the obvious way.

@@

In this section we discuss the notion of independence between two
consecutive rewriting steps, a fundamental ingredient of the rewriting
theory, which comes into play when viewing a sequence of steps as a
concurrent computation. We observe that moving from linear to
left-linear rules leads to the failure of various basic properties of
the classical notion of independence used in the linear setting and we
single out a framework in which these can be re-established, possibly
in a weakened form.

@@

Sequential independence is the canonical notion of independence in the DPO approach.
%recalled below.

@@

As we mentioned, for linear rewriting systems
%when viewing  derivations as concurrent computations, 
it is canonical to identify
derivations that are equal ``up to switching'', i.e. that differs
only for the order of independent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

Switch equivalence is now defined as the equivalence relating derivations that 
can be obtained one from another by a sequence of switches. Moreover, intermediate graphs can be taken up to isomorphism according to abstraction equivalence.

@@

The fact that sequential independence implies switchability always
holds for linear rules (see \Cref{prop:equi}). The result is so
indispensable that it is not even stated, in the sense that
in the literature switchability is not introduced axiomatically as in
Definition~\ref{def:switch}, but it is based on the explicit
construction of a switch.

@@

  Take the set $S:=\{a,b,c\}$ and endow it with the smallest order
  such that $a\leq b$ and $a\leq c$.  Take also the set $\mathbb{N}$
  ordered by $\geq$, so that $0$ is its maximum.  Let $\X$ be the
  category associated to the order
  $(S, \leq)\oplus (\mathbb{N}, \geq)$, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. Notice that the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a pushout
  because $b$ and $c$ do not have a supremum.  Consider a rewriting
  system whose set of rules $\R$ contains the following %two rules
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ of length $2$ below
  \[
    \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
      \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
      \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
      \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  for $i \in \{0,1\}$ and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback object (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

\begin{remark}
  By \Cref{thm:church} the existence of a strong independence pair
  entails switchability, which in turn entails weak sequential
  independence by construction. Strong Enforcing rewriting systems are exactly
  those rewriting systems in which these three notions coincide.
\end{remark}
%}

As a sanity check, note that a source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.
